Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-01-20 paper/sheaf_lics.tex
types + cocone + isSheaf (IsTrunc p)

But furthermore, one can defined a left adjoint to the inclusion, the
sheafification functor
%
$$
a_j : \E \rightarrow \Sh{\E}
$$
which exhibits $\Sh{\E}$ as a reflective
subcategory of $\E$ (which is a particular case of localization). This
means that logical principles valid in $\E$ are still valid in
$\Sh{\E}$.

@@

To extend Lawvere-Tierney sheafification to homotopy type theory, the
first thing to understand is that the construction can not be done in one
single step anymore. It must rather be performed by induction on the
level of homotopy types. More precisely, the first layer of
sheafification is defined for $\Type_{0}$, given a topology on
$\Type_{-1}$. This part corresponds to sheafification for toposes. Then,
assuming that the sheafification has been constructed up to level $n$,
one can defined the sheafification for $\Type_{n+1}$.
%
This inductive process in the definition of sheafification is similar
to construction in the Grothendieck setting, where the
sheafification is obtained by applying the plus-construction $n+1$
times to
$n$-types~\cite[Chapter~6]{lurie}.

This inductive step requires to formalize the notion of left-exact
reflective subuniverse, which corresponds to a stratified version of
left-exact modalities introduced in~\cite[Chapter~7]{hottbook}. A
Lawvere-Tierney topology can thus be seen as a left-exact modality on
$\Type_{-1}$ and sheafification as an inductive process that extends
it as a left-exact modality on $\Type_{n}$ for any $n$.

Compare to the definition of sheafification in topos theory, the main
change comes from the fact that in homotopy type theory, there is not
a single object classifier but rather an $n$-object classifier for
each $n\geq -2$ which classifies functions with $n$-truncated homotopy
fibers~\cite{sets_in_hott}. This means that when monomorphisms are used in topos-theoretic definition, the correct generalization may be $n$-truncated homotopy
fibers or embeddings (\ie $-1$-truncated homotopy
fibers) depending on the purpose of the definition.

@@

In this section, we review some basic definitions in homotopy type
theory that are central in our formalization but not specific to
sheafification. 
%
Section~\ref{ssec:hott} introduces the notion of homotopy types and
classifying objects as defined in~\cite{sets_in_hott}.
%
Section~\ref{sec:epi-mono-fact} presents the usual epi-mono
factorization system in the light of homotopy type theory.
%
Section~\ref{sec:colim-homot-type} introduces the notion of colimits
in homotopy type theory. 
%
And Section~\ref{sec:giraud-ax} defines a particular case of
Giraud-Rezk-Lurie axioms on higher toposes that we need in the
definition of sheafification.
%
The definition of Section~\ref{ssec:hott} and \ref{sec:epi-mono-fact}
are part of the Coq library
available at \url{https://github.com/HoTT/HoTT}, while other definitions
and theorems are specific to our formalization.

@@

\begin{itemize}
\item $\IsContr := \IsType {(-2)}$ and $\Contr := \Type_{-2}$
\item $\IsHProp := \IsType {(-1)}$ and $\HProp := \Type_{-1}$
\end{itemize}

@@

A map $f:A\to B$ is an \emph{embedding} when it is (-1)-truncated
(see definition in Section~\ref{ssec:hott}), and a \emph{surjection}
when it is (-1)-connected, that is when every fiber of $f$ is merely
inhabited (\ie $\|\fib f y\|$ holds for all $y$).

@@

Note that this generalizes to orthogonal factorization systems
constituted by $n$-truncated morphism and $n$-connected morphisms, for
every truncated level $n$. We will only use the factorization system
at level  $-1$ in the definition of sheafification.

@@

In the definition of (a special case of) Giraud-Rezk-Lurie axiom given in
Section~\ref{sec:giraud-ax}, the colimit of the {\em \v{C}ech nerve}
plays a central role.
% 
Following the notion of (homotopy) limits of diagrams over graphs that
has been defined in homotopy type theory~\cite{lumsdaine}, we present
here a definition of colimits of diagrams overs graphs. 
%
The main difference between limits and colimits is that limits are
simply given by $\Sigma$-types, and thus exists already in traditional
type theory, whereas the situation is more complicated for colimits as
it requires the use of higher inductive types.

@@

A colimit of a diagram $D$ over a graph $G$ is given by a type $P$
that defines a cone of $D$, plus the universal property that for any
type $X$, the canonical map that transforms a function $f : P
\rightarrow X$ to a cone of $D$ on $X$ is an isomorphism.
% 
\begin{defi}
Let $G$ be a graph, and $D$ be a diagram on $G$. 
Let $P:\Type$ together with
\begin{itemize}
\item a map $q_i : D_0(i) \to P$ for any
vertex $i:G_0$, \ie $$q : \prod_{i:G_0} D_0(i) \to P$$
\item for any vertices $i,j:G_0$ and all edges $\phi:G_1(i,j)$, a path
  $p_{i,j}^\phi : q_j \circ D_1(\phi) = q_i$, \ie
  $$p : \prod_{i,j:G_0} \prod_{\phi:G_1(i,j)} q_j \circ D_1(\phi) = q_i.$$
\end{itemize}

@@

Using higer inductive types, every diagram $D$ on a graph $G$ admits a
colimit in homotopy type thoery. This colimit $\colim D$ is given by
%
\begin{itemize}
\item a function $q:\prod_{i:G_0} D_0(i) \to \colim D$
\item for any $i,j:G_0$ and $\phi:G_1(i,j)$, a path
  $q_j \circ D_1(\phi) = q_i$.
\end{itemize}
%
Although we have proved in our formalization that $\colim D$ is
actually a colimit over $D$, we do not detail the proof here as the
existence of colimits is not used in the definition of the
sheafification process. We only make use of the following special case
of Giraud-Rezk-Lurie axioms.

@@

\begin{defi}
  Let $f:X \to Y$ be a map, and $p:\nat$. The $p$-pullback of $f$
  (also called the $p$-fold fiber product of $X$ over $Y$),
  noted $X\times_Y \cdots\times_Y X$, is
  the limit of the diagram
  $$\xymatrix{
    X \ar[rd]^f & \cdots & X \ar[ld]_f \\
      &    Y   &
  }$$
  with $p$ copies of $X$. The $0$-pullback of $f$ is $\Unit$.
\end{defi}

@@

The notion of left exact modalities plays a central role in this paper
because Lawvere-Tierney topologies corresponds exactly to left exact
modalities at level $-1$ and sheafification is described as an inductive
process that produces a left exact modality at level $n+1$ from a
level exact modality at level $n$.

Section~\ref{sec:definition} presents the definition and basic
properties of left exact modalities. 
%
Section~\ref{sec:new-type-theory} discusses how a left exact modality
induces a new homotopy theory, and in which case this theory is
consistent.
%
Finally, Section~\ref{sec:examples-left-exact} introduces some
example, and in particular the modality for the dense topology.

@@

As noticed in \cite{hottbook}, left exact modalities corresponds 
to sub-$(\infty,1)$-toposes. Indeed, modal types defines a new type
theory as modality preserves all type former, such as dependent
products or dependent sums.
% or inductive types (\nt{double check that}).

@@

A first example of left exact modality for every level $p$ is the {\em
  open modality of mere proposition $P$} defined as
$\modal_P T = P \to T$.
%
When one works in
the subuniverse defined by this modality, defining an inhabitant of a
type $T$ amounts to give an inhabitant of $P \to T$. 
%
Thus, working in this subuniverse amounts to add the axiom $P$ to the
theory. Of course, adding an axiom this way does not add any computational
content to $P$, and the consistency of the modallity corresponds to
the fact that $P \to \zero$ (or $\lnot P$) is not inhabited.

@@

The definition of a dense $n$-subobject requires more work as the
characteristic function $\chi_E$ of $E$ seen as an $n$-subobject of
itself is not the constant function equal to $\one$, as it is the case
for toposes, but rather the map 
%
$$\chi_E  = \lambda e . \ \sum_{e':E} e=e'.$$
%
This means that we need to take more care of the homotopy structure of
this charactestic function, and in particular, the definition of a
dense $n$-subobject comes with an additional coherence condition
(which is trivial when $n=-1$).

\begin{defi}
  Let $E$ be a type, and $\chi:E \to \Type_n$. The subobject $\iota :
  A \to E$ of $E$
  classified by $\chi$ is {\em dense} in $E$ when its $\modal_n$-closure
  is equvialent to $\chi_E$, \ie
  %
  $$\forall e:E,~ \left(\sum_{e':E} e=e'\right) \simeq \modal_n
  (\chi~e).$$
  % 
  and moreover, for any $x:A$, the following diagram 
  commutes, 
  $$\xymatrix{
    \sum_{e':A} x = e' \ar@{=}[r] \ar[d]_{\iota'} & (\chi \ x)
    \ar[d]^{\eta_{(\chi \ x)}}\\
    \sum_{e':E} x = e' \ar@{=}[r] & \modal_n (\chi \ x)
  }$$
  where $\iota': x \mapsto (\iota (\pi_1 x) ; \pi_2 x)$.
\end{defi}

@@

    It remains to prove that for every $(u,x)$ inhabiting the
    $\Sigma$-type above, $u$ is in $T\to\Type_n^\modal$, \ie that for
    every $t:T$, $\IsType n (u\, t)$.  But for any truncation index $p$,
    the predicate $\IsType p:\HProp$ is a sheaf, so we can get rid
    of $\modal_{-1}$ and of the truncation, which tells us that for
    every 
    $t:T$, $u\, t = \modal_n(a=t) : \Type_n$.
  \end{itemize}
\end{proof}

@@

More importantly, there are few axioms in the formalization that are
due to an unsatisfactory management of polymorphic universe levels in
Coq.
%
Universes are used in type theory to ensure consistency by checking
that definitions are well-stratified according to a certain hierarchy.
%
Universe polymorphism~\cite{sozeau2014universe} supports generic
definitions over universes, reusable at different levels.
%
Although the presence of universe polymorphism is mandatory for our
formalization, its implementation is still to rigid to allow a
complete formalization of our work for the following reasons.

>>>>

Furthermore, one can define a left adjoint to the inclusion, the
sheafification functor
%
$$
a_j : \E \rightarrow \Sh{\E}
$$
which exhibits $\Sh{\E}$ as a reflective
subcategory of $\E$ (which is a particular case of localization). This
means that logical principles valid in $\E$ are still valid in
$\Sh{\E}$.

@@

To extend Lawvere-Tierney sheafification to homotopy type theory, the
first thing to understand is that the construction can not be done in one
single step anymore. It must rather be performed by induction on the
level of homotopy types. More precisely, the first layer of
sheafification is defined for $\Type_{0}$, given a topology on
$\Type_{-1}$. This part corresponds to sheafification for toposes. Then,
assuming that the sheafification has been constructed up to level $n$,
one can define the sheafification for $\Type_{n+1}$.
%
This inductive process in the definition of sheafification is similar
to construction in the Grothendieck setting, where the
sheafification is obtained by applying the plus-construction $n+1$
times to
$n$-types~\cite[Chapter~6]{lurie}.

This inductive step requires to formalize the notion of left-exact
reflective subuniverse, which corresponds to a stratified version of
left-exact modality introduced in~\cite[Chapter~7]{hottbook}. A
Lawvere-Tierney topology can thus be seen as a left-exact modality on
$\Type_{-1}$ and sheafification as an inductive process that extends
it as a left-exact modality on $\Type_{n}$ for any $n$.

Compare to the definition of sheafification in topos theory, the main
change comes from the fact that in homotopy type theory, there is not
a single object classifier but rather an $n$-object classifier for
each $n\geq -2$ which classifies functions with $n$-truncated homotopy
fibers~\cite{sets_in_hott}. This means that when monomorphisms are used in topos-theoretic definition, the correct generalization may be $n$-truncated homotopy
fibers or embeddings (\ie $(-1)$-truncated homotopy
fibers) depending on the purpose of the definition.

@@

In this section, we review some basic definitions in homotopy type
theory that are central in our formalization but not specific to
sheafification. 
%
Section~\ref{ssec:hott} introduces the notion of homotopy types and
classifying objects as defined in~\cite{sets_in_hott}.
%
Section~\ref{sec:epi-mono-fact} presents the usual epi-mono
factorization system in the light of homotopy type theory.
%
Section~\ref{sec:colim-homot-type} introduces the notion of colimits
in homotopy type theory. 
%
And Section~\ref{sec:giraud-ax} defines a particular case of
Giraud-Rezk-Lurie axioms on higher toposes that we need in the
definition of sheafification.
%
The definitions of Section~\ref{ssec:hott} and \ref{sec:epi-mono-fact}
are part of the Coq library
available at \url{https://github.com/HoTT/HoTT}, while other definitions
and theorems are specific to our formalization.

@@

\begin{itemize}
\item $\IsContr \defeq \IsType {(-2)}$ and $\Contr \defeq \Type_{-2}$
\item $\IsHProp \defeq \IsType {(-1)}$ and $\HProp \defeq \Type_{-1}$
\end{itemize}

@@

A map $f:A\to B$ is an \emph{embedding} when it is $(-1)$-truncated
(see definition in Section~\ref{ssec:hott}), and a \emph{surjection}
when it is $(-1)$-connected, that is when every fiber of $f$ is merely
inhabited (\ie $\|\fib f y\|$ holds for all $y$).

@@

Note that this generalizes to orthogonal factorization systems
constituted by $n$-truncated morphisms and $n$-connected morphisms, for
every truncated level $n$. We will only use the factorization system
at level  $-1$ in the definition of sheafification.

@@

In the definition of (a special case of) Giraud-Rezk-Lurie axiom given in
Section~\ref{sec:giraud-ax}, the colimit of the {\em \v{C}ech nerve}
plays a central role.
% 
Following the notion of (homotopy) limits of diagrams over graphs that
has been defined in homotopy type theory~\cite{lumsdaine}, we present
here a definition of colimits of diagrams overs graphs. 
%
The main difference between limits and colimits is that limits are
simply given by $\Sigma$-types, and thus exist already in traditional
type theory, whereas the situation is more complicated for colimits as
it requires the use of higher inductive types.

@@

A colimit of a diagram $D$ over a graph $G$ is given by a type $P$
that defines a cocone of $D$, plus the universal property that for any
type $X$, the canonical map that transforms a function $f : P
\rightarrow X$ to a cocone of $D$ on $X$ is an isomorphism.
% 
\begin{defi}
Let $G$ be a graph, and $D$ be a diagram on $G$. 
Let $P:\Type$ together with
\begin{itemize}
\item a map $q_i : D_0(i) \to P$ for any
vertex $i:G_0$, \ie $$q : \prod_{i:G_0} D_0(i) \to P$$
\item for any vertices $i,j:G_0$ and all edges $\phi:G_1(i,j)$, a path
  $p_{i,j}^\phi : q_j \circ D_1(\phi) = q_i$, \ie
  $$p : \prod_{i,j:G_0} \prod_{\phi:G_1(i,j)} q_j \circ D_1(\phi) = q_i.$$
\end{itemize}

@@

Using higher inductive types, every diagram $D$ on a graph $G$ admits a
colimit in homotopy type theory. This colimit $\colim D$ is given by
%
\begin{itemize}
\item a function $q:\prod_{i:G_0} D_0(i) \to \colim D$
\item for any $i,j:G_0$ and $\phi:G_1(i,j)$, a path
  $q_j \circ D_1(\phi) = q_i$.
\end{itemize}
%
Although we have proved in our formalization that $\colim D$ is
actually a colimit over $D$, we do not detail the proof here as the
existence of colimits is not used in the definition of the
sheafification process. We only make use of the following special case
of Giraud-Rezk-Lurie axioms.

@@

\begin{defi}
  Let $f:X \to Y$ be a map, and $p:\nat$. The $p$-pullback of $f$
  (also called the $p$-fold fiber product of $X$ over $Y$),
  noted $X\times_Y \cdots\times_Y X$, is
  the limit of the diagram
  $$\xymatrix{
    X \ar[rd]^f & \cdots & X \ar[ld]_f \\
      &    Y   &
  }$$
  with $p$ copies of $X$. The $0$-pullback of $f$ is $\one$.
\end{defi}

@@

The notion of left exact modalities plays a central role in this paper
because Lawvere-Tierney topologies correspond exactly to left exact
modalities at level $-1$ and sheafification is described as an inductive
process that produces a left exact modality at level $n+1$ from a
level exact modality at level $n$.

Section~\ref{sec:definition} presents the definition and basic
properties of left exact modalities. 
%
Section~\ref{sec:new-type-theory} discusses how a left exact modality
induces a new homotopy theory, and in which case this theory is
consistent.
%
Finally, Section~\ref{sec:examples-left-exact} introduces some
examples, and in particular the modality for the dense topology.

@@

As noticed in \cite{hottbook}, left exact modalities correspond
to sub-$(\infty,1)$-toposes. Indeed, modal types defines a new type
theory as modality preserves all type former, such as dependent
products or dependent sums.
% or inductive types (\nt{double check that}).

@@

A first example of left exact modality for every level $p$ is the {\em
  open modality of mere proposition $P$} defined as
$\modal_P T = P \to T$.
%
When one works in
the subuniverse defined by this modality, defining an inhabitant of a
type $T$ amounts to give an inhabitant of $P \to T$. 
%
Thus, working in this subuniverse amounts to add the axiom $P$ to the
theory. Of course, adding an axiom this way does not add any computational
content to $P$, and the consistency of the modality corresponds to
the fact that $P \to \zero$ (or $\lnot P$) is not inhabited.

@@

The definition of a dense $n$-subobject requires more work as the
characteristic function $\chi_E$ of $E$ seen as an $n$-subobject of
itself is not the constant function equal to $\one$, as it is the case
for toposes, but rather the map 
%
$$\chi_E  = \lambda e . \ \sum_{e':E} e=e'.$$
%
This means that we need to take more care of the homotopy structure of
this characteristic function, and in particular, the definition of a
dense $n$-subobject comes with an additional coherence condition
(which is trivial when $n=-1$).

\begin{defi}
  Let $E$ be a type, and $\chi:E \to \Type_n$. The subobject $\iota :
  A \to E$ of $E$
  classified by $\chi$ is {\em dense} in $E$ when its $\modal_n$-closure
  is equivalent to $\chi_E$, \ie
  %
  $$\forall e:E,~ \left(\sum_{e':E} e=e'\right) \simeq \modal_n
  (\chi~e).$$
  % 
  and moreover, for any $x:A$, the following diagram 
  commutes, 
  $$\xymatrix{
    \sum_{e':A} x = e' \ar@{=}[r] \ar[d]_{\iota'} & (\chi \ x)
    \ar[d]^{\eta_{(\chi \ x)}}\\
    \sum_{e':E} x = e' \ar@{=}[r] & \modal_n (\chi \ x)
  }$$
  where $\iota': x \mapsto (\iota (\pi_1 x) ; \pi_2 x)$.
\end{defi}

@@

    It remains to prove that for every $(u,x)$ inhabiting the
    $\Sigma$-type above, $u$ is in $T\to\Type_n^\modal$, \ie that for
    every $t:T$, $\IsType n (u\, t)$.  But for any truncation index
    $p$,
    the type $\IsType p X:\HProp$ is a sheaf as soon as $X$ is, so we can get rid
    of $\modal_{-1}$ and of the truncation, which tells us that for
    every 
    $t:T$, $u\, t = \modal_n(a=t) : \Type_n$.
  \end{itemize}
\end{proof}

@@

More importantly, there are few axioms in the formalization that are
due to an unsatisfactory management of polymorphic universe levels in
Coq.
%
Universes are used in type theory to ensure consistency by checking
that definitions are well-stratified according to a certain hierarchy.
%
Universe polymorphism~\cite{sozeau2014universe} supports generic
definitions over universes, reusable at different levels.
%
Although the presence of universe polymorphism is mandatory for our
formalization, its implementation is still too rigid to allow a
complete formalization of our work for the following reasons.
